# Accidentes y PSAT scores
## R notebook for reproducibility
### Wilfrido J. Paredes

A sample of the data
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(plotly)

accidentes <- read_excel('accidentes.xlsx')
head(accidentes)
```
We convert some numerical variables to categorical in order to work in proper way. For example, the type of the accident.
Also, we defined new variables:

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes <- accidentes %>% mutate('Magnitude' = AUTOMOVIL+CAMPASAJ+MICROBUS+PASCAMION+OMNIBUS+TRANVIA+CAMIONETA+CAMION+TRACTOR+FERROCARRI+MOTOCICLET+BICICLETA+OTROVEHIC)
accidentes <- accidentes %>% mutate('Severity' = CONDMUERTO+PASAMUERTO+PEATMUERTO+CICLMUERTO+OTROMUERTO+0.5*(CONDHERIDO+PASAHERIDO+PEATHERIDO+CICLHERIDO+OTROHERIDO))
accidentes <- accidentes %>% mutate('Deaths' = CONDMUERTO+PASAMUERTO+PEATMUERTO+CICLMUERTO+OTROMUERTO)
accidentes <- accidentes %>% mutate("Is_Deaths" = if_else(Deaths > 0, "Involved", "Not involved"))
accidentes <- accidentes %>% mutate('Injured' = CONDHERIDO+PASAHERIDO+PEATHERIDO+CICLHERIDO+OTROHERIDO)
accidentes <- accidentes %>% mutate("Is_Injured" = if_else(Injured > 0, "Involved", "Not involved"))
accidentes <- accidentes %>% mutate('Rate DI' = if_else(Injured > 0, if_else(Deaths > 0, Deaths/Injured,0),if_else(Deaths > 0, 10, 0)))
accidentes <- accidentes %>% mutate('Pedestrian' = if_else(PEATMUERTO+PEATHERIDO > 0, 'Involved', 'Not involved'))
accidentes <- accidentes %>% mutate('Cyclist' = if_else(CICLMUERTO+CICLHERIDO+BICICLETA > 0, 'Involved', 'Not involved'))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes$TIPACCID <- as.factor(accidentes$TIPACCID)
```

#### Accident indicators 2019 vs 2020
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes_dummy <- accidentes %>% group_by(ANIO) %>% summarise(
    'Magnitude Median' = median(Magnitude),
    'Severity Median' = median(Severity),
    'Deaths Median' = median(`Rate DI`),
    'Rate of pedestrians involved' = sum(Pedestrian=='Involved')/n()*100,
    'Rate of cyclists involved' = sum(Cyclist=='Involved')/n()*100
)

print(accidentes_dummy)
m2019 <- accidentes %>% filter(ANIO==2019) %>% pull(Magnitude)
m2020 <- accidentes %>% filter(ANIO==2020) %>% pull(Magnitude)

figm <- plot_ly(alpha = 0.6)
figm <- figm %>% add_histogram(x = ~m2019)
figm <- figm %>% add_histogram(x = ~m2020)
figm <- figm %>% layout(barmode = "overlay")

s2019 <- accidentes %>% filter(ANIO==2019) %>% pull(Severity)
s2020 <- accidentes %>% filter(ANIO==2020) %>% pull(Severity)

figs <- plot_ly(alpha = 0.6)
figs <- figs %>% add_histogram(x = ~s2019)
figs <- figs %>% add_histogram(x = ~s2020)
figs <- figs %>% layout(barmode = "overlay")

fig <- subplot(figm, figs)
fig
```

In general, there is no change in accident dynamics. Accidents generally involve the same magnitude, severity, pedestrian injuries, cyclists involved in both years. This means that the pandemic did not affect this.

#### Type of the accidents 2019 vs 2020
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes_por_calle <- accidentes %>% select(Cve_Segmen,ANIO,TIPACCID) %>%
    group_by(Cve_Segmen,ANIO,TIPACCID) %>%
    mutate(Counts = n()) %>% ungroup() %>% distinct() %>%
    spread(key=TIPACCID, value = Counts, fill = 0) %>% 
    select(Cve_Segmen,ANIO,`1`,`2`,`3`,`4`,`5`,`6`,`7`,`8`,`9`,`10`,`11`,`12`) %>%
    distinct()

accidentes2019 <- log(colSums(accidentes_por_calle %>% filter(ANIO == 2019) %>% select(-Cve_Segmen,-ANIO)))
accidentes2020 <- log(colSums(accidentes_por_calle %>% filter(ANIO == 2020) %>% select(-Cve_Segmen,-ANIO)))
accidentes2020[9] <- 0

aux <- data.frame(accidentes2019,accidentes2020)
tipo_accidentes <- as.data.frame(t(aux))

fig <- plot_ly(tipo_accidentes, x=c(2019,2020), y = ~`1`, name='1', type='scatter', mode='lines')
fig <- fig %>% add_trace( y = ~`2`, name='2')
fig <- fig %>% add_trace( y = ~`3`, name='3')
fig <- fig %>% add_trace( y = ~`4`, name='4')
fig <- fig %>% add_trace( y = ~`5`, name='5')
fig <- fig %>% add_trace( y = ~`6`, name='6')
fig <- fig %>% add_trace( y = ~`7`, name='7')
fig <- fig %>% add_trace( y = ~`8`, name='8')
fig <- fig %>% add_trace( y = ~`9`, name='9')
fig <- fig %>% add_trace( y = ~`10`, name='10')
fig <- fig %>% add_trace( y = ~`11`, name='11')
fig <- fig %>% add_trace( y = ~`12`, name='12')

fig <- fig %>% layout(
    xaxis = list(range = c(2017,2022), tickvals = list(2019,2020), tickmode = "array"),
    yaxis = list(showgrid=FALSE, range = c(0,9), tickvals = list(2019,2020), tickmode = "array", title='Number of cases'))

fig <- fig %>% add_annotations(x = rep(2020+0.1,12),
                  y = accidentes2020,
                  text = as.character(round(exp(accidentes2020),0)),
                  xref = "x",
                  yref = "y",
                  showarrow = FALSE,
                  ax = 20,
                  ay = -40)

fig <- fig %>% add_annotations(x = rep(2019-0.1,12),
                  y = accidentes2019,
                  text = as.character(round(exp(accidentes2019),0)),
                  xref = "x",
                  yref = "y",
                  showarrow = FALSE,
                  ax = 20,
                  ay = -40)
fig
```
The figure shows that there is a general decrease in accidents. In particular, type 11 accidents decreased by 37.86%. The decrease is due to the decrease in mobility due to the pandemic. However, it is necessary to emphasize that the indicators remained the same. In other words, accidents are still as serious, but occurred in lesser numbers. 

#### Bicycle accidents
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes_bicicleta <- accidentes %>% select(Cve_Segmen,ANIO,Cyclist,longitud,I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL) %>%
    group_by(Cve_Segmen,ANIO,Cyclist) %>%
    mutate(Counts = n()) %>% 
    group_by(Cve_Segmen,ANIO) %>% 
    mutate(I_BANQ = mean(I_BANQ_VAL),
        I_ARBOL = mean(I_ARBOL_VA),
        I_RAMP = mean(I_RAMP_VAL),
        I_ALUM = mean(I_ALUM_VAL),
        I_SENA = mean(I_SENA_VA),
        A_DENS = mean(A_DENS_VAL),
        A_ESCU = mean(A_ESCU_VAL),
        A_COME = mean(A_COME_VAL),
        A_DEHA = mean(A_DEHA_VAL),
        T_CEPO = mean(T_CEPO_VAL),
        T_JERA = mean(T_JERA_VAL),
        T_TRPU = mean(T_TRPU_VAL),
        T_BRT = mean(T_BRT_VAL),
        T_METR = mean(T_METR_VAL)) %>% 
    ungroup() %>% 
    select(-c(I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL)) %>% 
    distinct() %>%
    spread(key=Cyclist, value = Counts, fill = 0)

accidentes_bicicleta <- accidentes_bicicleta %>% mutate('Rate of involved cyclists' = Involved/(Involved+`Not involved`))
accidentes_bicicleta <- accidentes_bicicleta %>% mutate('Invol' = if_else(Involved > 0, 1, 0))
```

```{r, echo=FALSE}
ss <- accidentes_bicicleta$ANIO == 2019
model_bici <- glm(formula = Involved ~ I_BANQ+T_JERA, data = accidentes_bicicleta, subset = ss, family = poisson)
summary(model_bici)
```
There is no appreciable feature that favors accidents involving cyclists. In some cases, graphically it can be seen that there is; however, the numerical summaries of the data indicate that it is only due to the low diversity of observations in this variable.


#### Injured accidents
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes_injured <- accidentes %>% select(Cve_Segmen,ANIO,Is_Injured,longitud,I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL) %>%
    group_by(Cve_Segmen,ANIO,Is_Injured) %>%
    mutate(Counts = n()) %>% 
    group_by(Cve_Segmen,ANIO) %>% 
    mutate(I_BANQ = mean(I_BANQ_VAL),
        I_ARBOL = mean(I_ARBOL_VA),
        I_RAMP = mean(I_RAMP_VAL),
        I_ALUM = mean(I_ALUM_VAL),
        I_SENA = mean(I_SENA_VA),
        A_DENS = mean(A_DENS_VAL),
        A_ESCU = mean(A_ESCU_VAL),
        A_COME = mean(A_COME_VAL),
        A_DEHA = mean(A_DEHA_VAL),
        T_CEPO = mean(T_CEPO_VAL),
        T_JERA = mean(T_JERA_VAL),
        T_TRPU = mean(T_TRPU_VAL),
        T_BRT = mean(T_BRT_VAL),
        T_METR = mean(T_METR_VAL)) %>% 
    ungroup() %>% 
    select(-c(I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL)) %>% 
    distinct() %>%
    spread(key=Is_Injured, value = Counts, fill = 0)

accidentes_injured <- accidentes_injured %>% mutate('Rate of injured' = Involved/(Involved+`Not involved`))
accidentes_injured <- accidentes_injured %>% mutate('Invol' = if_else(Involved > 0, 1, 0))
```

```{r, echo=FALSE}
ss <- accidentes_injured$ANIO == 2019
model_injured <- glm(formula = Involved ~ I_BANQ+I_ARBOL+I_RAMP+I_ALUM+I_SENA+A_DENS+A_ESCU+A_COME+A_DEHA+T_CEPO+T_JERA+T_TRPU, data = accidentes_injured, subset = ss, family = poisson)
summary(model_injured)

model_injured <- glm(formula = Involved ~ I_BANQ+I_RAMP+T_CEPO+T_JERA+T_TRPU, data = accidentes_injured, subset = ss, family = poisson)
summary(model_injured)

```


#### Injured accidents
```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
accidentes_death <- accidentes %>% select(Cve_Segmen,ANIO,Is_Deaths,longitud,I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL) %>%
    group_by(Cve_Segmen,ANIO,Is_Deaths) %>%
    mutate(Counts = n()) %>% 
    group_by(Cve_Segmen,ANIO) %>% 
    mutate(I_BANQ = mean(I_BANQ_VAL),
        I_ARBOL = mean(I_ARBOL_VA),
        I_RAMP = mean(I_RAMP_VAL),
        I_ALUM = mean(I_ALUM_VAL),
        I_SENA = mean(I_SENA_VA),
        A_DENS = mean(A_DENS_VAL),
        A_ESCU = mean(A_ESCU_VAL),
        A_COME = mean(A_COME_VAL),
        A_DEHA = mean(A_DEHA_VAL),
        T_CEPO = mean(T_CEPO_VAL),
        T_JERA = mean(T_JERA_VAL),
        T_TRPU = mean(T_TRPU_VAL),
        T_BRT = mean(T_BRT_VAL),
        T_METR = mean(T_METR_VAL)) %>% 
    ungroup() %>% 
    select(-c(I_BANQ_VAL,I_ARBOL_VA,I_RAMP_VAL,I_ALUM_VAL,I_SENA_VA,A_DENS_VAL,A_ESCU_VAL,A_COME_VAL,A_DEHA_VAL,T_CEPO_VAL,T_JERA_VAL,T_TRPU_VAL,T_BRT_VAL,T_METR_VAL)) %>% 
    distinct() %>%
    spread(key=Is_Deaths, value = Counts, fill = 0)

accidentes_death <- accidentes_death %>% mutate('Rate of deaths' = Involved/(Involved+`Not involved`))
accidentes_death <- accidentes_death %>% mutate('Invol' = if_else(Involved > 0, 1, 0))
```

```{r, echo=FALSE}
ss <- accidentes_death$ANIO == 2019
model_death <- glm(formula = Involved ~ I_BANQ+I_ARBOL+I_RAMP+I_ALUM+I_SENA+A_DENS+A_ESCU+A_COME+A_DEHA+T_CEPO+T_JERA+T_TRPU, data = accidentes_death, subset = ss, family = poisson)
summary(model_death)

model_death <- glm(formula = Involved ~ I_SENA+A_DEHA+T_JERA, data = accidentes_death, subset = ss, family = poisson)
summary(model_death)

```


```{r, echo=FALSE}
accidentes_bicicleta_year <- accidentes_bicicleta %>% filter(ANIO == 2019)
cutpoint <- quantile(accidentes_bicicleta_year$`Rate of involved cyclists`, 0.95)
accidentes_bicicleta_year %>% filter(`Rate of involved cyclists` > cutpoint)

pl_colorscale = list(c(0.0, '#119dff'),
                  c(0.5, '#119dff'),
                  c(0.5, '#ef553b'),
                  c(1, '#ef553b'))

axis = list(showline=FALSE,
            zeroline=FALSE,
            gridcolor='#ffff',
            ticklen=4,
            titlefont=list(size=13))

fig <- accidentes_bicicleta_year %>% plot_ly()
fig <- fig %>%
  add_trace(
    type = 'splom',
    dimensions = list(
      list(label='Sidewalk', values=~I_BANQ),
      list(label='Tree', values=~I_ARBOL),
      list(label='Slope', values=~I_RAMP),
      list(label='Street lighting', values=~I_ALUM),
      list(label='Traffic sign', values=~I_SENA),
      list(label='Dens', values=~A_DENS),
      list(label='Schools', values=~A_ESCU),
      list(label='Stores', values=~A_COME),
      list(label='Deha', values=~A_DEHA),
      list(label='Cepo', values=~T_CEPO),
      list(label='Jera', values=~T_JERA),
      list(label='Trpu', values=~T_TRPU),
      list(label='Brt', values=~T_BRT),
      list(label='Metr', values=~T_METR),
      list(label='Rate', values=~`Rate of involved cyclists`)
    ),

    text=~factor(Invol, labels=c("Not involved","Cyclist involved")),
    diagonal=list(visible=FALSE),
    marker = list(
      color = ~Invol,
      colorscale = pl_colorscale,
      size = 5,
      line = list(
        width = 1,
        color = 'rgb(230,230,230)'
      )
    )
  ) 

fig <- fig %>% layout(
    title = "Scatterplot Matrix (SPLOM) for biclycle accidents",
    hovermode='closest',
    dragmode = 'select',
    plot_bgcolor='rgba(240,240,240, 0.95)',
    xaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4, titlefont=list(size=13)),
    yaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4, titlefont=list(size=13)),
    xaxis2=axis,
    xaxis3=axis,
    xaxis4=axis,
    xaxis5=axis,
    xaxis6=axis,
    xaxis7=axis,
    xaxis8=axis,
    xaxis9=axis,
    xaxis10=axis,
    xaxis11=axis,
    xaxis12=axis,
    xaxis13=axis,
    xaxis14=axis,
    xaxis15=axis,
    yaxis2=axis,
    yaxis3=axis,
    yaxis4=axis,
    yaxis5=axis,
    yaxis6=axis,
    yaxis7=axis,
    yaxis8=axis,
    yaxis9=axis,
    yaxis10=axis,
    yaxis11=axis,
    yaxis12=axis,
    yaxis13=axis,
    yaxis14=axis,
    yaxis15=axis,
  )

summary(accidentes_bicicleta %>% filter(T_JERA==18, ANIO==2019) %>% select(Involved))
fig

```

```{r, echo=FALSE}
head(accidentes_bicicleta_year)
cor(accidentes_bicicleta_year %>% select(-c(Cve_Segmen,ANIO,Involved,`Not involved`,Invol)))
```